Targeted interrogation of mutational hotspots in rheumatoid arthritis

### 2024-03-06 infercnv/reports/gains_and_losses.Rmd
add chisq test, fix chr gain contingency plotting

```{r chr7_and_chr5}
events <-
  sce@colData %>%
  tibble::as_tibble() %>%
  dplyr::select(sample, cell, cell_group, 
                dplyr::ends_with('_gain'), 
                dplyr::ends_with('_loss')) %>%
  tidyr::pivot_longer(-c('sample', 'cell', 'cell_group')) %>%
  dplyr::group_by(sample, cell, cell_group) %>%
  # remove cells in which no calls made
  dplyr::filter(!all(is.na(value))) %>%
  dplyr::mutate(
    event = dplyr::case_when(
      all(value == 0 | is.na(value)) ~ 'WT',
      value == 1 ~ name,
      TRUE ~ NA)) %>%
  dplyr::summarise(
    events = paste(sort(unique(na.omit(event))), collapse = ','),
    n_events = sum(value, na.rm = T))

p_dat <-
  events %>%
  dplyr::ungroup() %>%
  dplyr::add_count(events) %>%
  dplyr::mutate(prop = n / dplyr::n()) %>%
  # ignore events that happen in <10 cells
  dplyr::filter(events != 'WT', n >= 10) %>%
  dplyr::mutate(
    events = events %>% 
      gsub('chr', '', .) %>% 
      gsub('_gain', '+', .) %>%
      gsub('_loss', '+', .),
    events = forcats::fct_infreq(events),
    label1 = scales::label_comma()(n), 
    label = dplyr::case_when(
      n < 10 ~ label1,
      TRUE ~ paste0(label1, ' (', round(prop*100, 2), '%)')
    ))
p_dat %>%
  ggplot(aes(x = events, fill = cell_group, label = label)) +
  geom_bar() +
  geom_text(aes(y = n), hjust = 0, vjust = -0.1, angle = 45, size = 2) +
  scale_x_discrete(guide = guide_axis(angle = -90)) +
  ylim(c(0, max(p_dat$n) * 1.1)) +
  theme_classic() +
  theme(legend.position = 'bottom')
p_dat %>%
  ggplot(aes(x = forcats::fct_infreq(events), fill = sample, label = label)) +
  geom_bar() +
  geom_text(aes(y = n), hjust = 0, vjust = -0.1, angle = 45, size = 2) +
  scale_x_discrete(guide = guide_axis(angle = -90)) +
  ylim(c(0, max(p_dat$n) * 1.1)) +
  theme_classic() +
  theme(legend.position = 'none')
```

>>>>

```{r chr7_and_others}
events <-
  sce@colData %>%
  tibble::as_tibble() %>%
  dplyr::select(sample, cell, cell_group, 
                dplyr::ends_with('_gain'), 
                dplyr::ends_with('_loss')) %>%
  tidyr::pivot_longer(-c('sample', 'cell', 'cell_group')) %>%
  dplyr::group_by(sample, cell, cell_group) %>%
  # remove cells in which no calls made
  dplyr::filter(!all(is.na(value))) %>%
  dplyr::mutate(
    event = dplyr::case_when(
      all(value == 0 | is.na(value)) ~ 'WT',
      value == 1 ~ name,
      TRUE ~ NA)) %>%
  dplyr::summarise(
    events = paste(sort(unique(na.omit(event))), collapse = ','),
    n_events = sum(value, na.rm = T))

p_dat <-
  events %>%
  dplyr::ungroup() %>%
  dplyr::add_count(events) %>%
  dplyr::mutate(prop = n / dplyr::n()) %>%
  # ignore events that happen in <= 25 cells
  dplyr::filter(events != 'WT', n >= 25) %>%
  dplyr::mutate(
    events = events %>% 
      gsub('chr', '', .) %>% 
      gsub('_gain', '+', .) %>%
      gsub('_loss', '-', .),
    events = forcats::fct_infreq(events),
    label1 = scales::label_comma()(n), 
    label = dplyr::case_when(
      n < 10 ~ label1,
      TRUE ~ paste0(label1, ' (', round(prop*100, 2), '%)')
    ))

# plot
p <-
  p_dat %>%
  ggplot(aes(x = events)) +
  geom_text(aes(y = n + 50,  label = label), 
            vjust = 0.5, hjust = 0, angle = 90, size = 2) +
  scale_x_discrete(guide = guide_axis(angle = -90)) +
  ylim(c(0, max(p_dat$n) * 1.1)) +
  theme_classic() +
  theme(legend.position = 'bottom')
p + 
  geom_bar(aes(fill = sample))
p + 
  geom_bar(aes(fill = cell_group)) +
  theme(legend.position = 'none')
```

We also can run a Fisher exact test to test whether chr7+ and chr5+ gains co-occur more often than would be expected by chance, and also test all other co-occurences.

```{r cont_tab}
cont <-
  sce@colData %>%
  tibble::as_tibble() %>%
  dplyr::mutate(chr7 = ifelse(chr7_gain == 1, 'chr7+', 'chr7_WT'),
                chr5 = ifelse(chr5_gain == 1, 'chr5+', 'chr5_WT')) 
table(cont$chr7, cont$chr5) %>%
  knitr::kable()
```

```{r fisher_exact}
chisq <- xfun::cache_rds({

  chisq <-
    colnames(sce@colData)[grepl('_gain$|_loss$', colnames(sce@colData))] %>%
    purrr::set_names() %>%
    purrr::map(function(event) {

      if (1 %in% cont[, event, drop = T]) {

        # run chi-squared test
        chisq <- chisq.test(sce@colData[, 'chr7_gain', drop = T], 
                            sce@colData[, event, drop = T])

         # return relevant stats
        tibble::tibble(
          X_squared = chisq$statistic, 
          p = chisq$p.value, 
          df = chisq$parameter['df'])

      }

    }) %>%
    dplyr::bind_rows(.id = 'event') %>%
    # remove chr7+
    dplyr::filter(event != 'chr7_gain') %>%
    dplyr::arrange(p)

  # return
  chisq

}, file = 'chisq.rds', rerun = params$rerun)

chisq %>%
  knitr::kable()
```

### 2024-03-06 scomatic/reports/targeted_mutation_calling.Rmd
add chisq test, fix chr gain contingency plotting

```{r mut_hotspots}
# read in bam mappings, get celltype-split bams only
mappings <-
  readr::read_csv(params$bam_mappings) %>%
  dplyr::filter(!is.na(celltype))

@@

# add refs, compare
muts <-
  sites %>%
  dplyr::left_join(refs) %>%
  tidyr::pivot_longer(cols = dplyr::all_of(nucleotides),
                      values_to = 'nt_n_reads', names_to = 'nt') %>%
  # # ignore indels for now
  # dplyr::filter(!nt %in% c('-', '_', 'INS', 'ins')) %>%
  # get number of ref and non-ref alleles at each position
  dplyr::mutate(n_ref = ifelse(ref == toupper(nt), nt_n_reads, 0),
                n_non_ref = ifelse(ref != toupper(nt), nt_n_reads, 0)) %>%
  dplyr::group_by(id, celltype, chr, pos, mut_pos, gene, total, ref) %>%
  dplyr::summarise(n_ref = sum(n_ref), n_non_ref = sum(n_non_ref)) %>%
  dplyr::mutate(vaf = n_non_ref / total)

# plot positions
unique(sites$gene) %>%
  purrr::walk(function(i) {
    p <-
      sites %>%
      dplyr::filter(gene == i) %>%
      dplyr::left_join(refs, by = c('chr', 'pos')) %>%
      tidyr::pivot_longer(cols = dplyr::all_of(nucleotides),
                          values_to = 'nt_n_reads', names_to = 'nt') %>%
      dplyr::mutate(
        strandless_nt = 
          ifelse(nt %in% c('-', '_'), 'DEL', toupper(nt)),
        is_ref = ifelse(strandless_nt == ref, 'ref', 'alt'),
        strandless_nt = paste0(strandless_nt, '_', is_ref) %>% 
          forcats::fct_relevel(unique(c(paste0(topupper(nucleotides), '_ref'),
                                        paste0(toupper(nucleotides), '_alt'))))) %>%
      ggplot(aes(x = pos, y = nt_n_reads, fill = strandless_nt)) + 
      geom_vline(aes(xintercept = mut_pos)) +
      geom_col(position = 'stack') +
      theme_bw() +
      theme(axis.text.x = element_blank()) +
      ggh4x::facet_grid2(celltype ~ mut_pos, scales = 'free_x') +
      scale_fill_manual(values = c(A_ref = 'darkgreen', A_alt = 'lightgreen',
                                   C_ref = 'darkblue', C_alt = 'lightblue',
                                   G_ref = 'darkorange3', G_alt = 'peachpuff',
                                   T_ref = 'darkred', T_alt = 'salmon',
                                   INS_alt = 'black', DEL_alt = 'grey')) +
      ggtitle(i)
    print(p)
    })
```

>>>>

```{r mut_hotspots}
# read in bam mappings, get celltype-split bams only
mappings <-
  readr::read_csv(params$bam_mappings) %>%
  dplyr::filter(!is.na(celltype)) %>%
  # temporarily remove problematic BRI-401 endo bam
  dplyr::filter(!(id == 'BRI_401' & celltype == 'Endothelial_cell'))

@@

# add refs, compare
muts <-
  sites %>%
  dplyr::left_join(refs) %>%
  tidyr::pivot_longer(cols = dplyr::all_of(nucleotides),
                      values_to = 'nt_n_reads', names_to = 'nt') %>%
  # get number of ref and non-ref alleles at each position
  dplyr::mutate(n_ref = ifelse(ref == toupper(nt), nt_n_reads, 0),
                n_non_ref = ifelse(ref != toupper(nt), nt_n_reads, 0)) %>%
  dplyr::group_by(id, celltype, chr, pos, mut_pos, gene, total, ref) %>%
  dplyr::summarise(n_ref = sum(n_ref), n_non_ref = sum(n_non_ref)) %>%
  dplyr::mutate(vaf = n_non_ref / total)
```

```{r plot}
# plot positions
unique(sites$gene) %>%
  purrr::walk(function(i) {
    p <-
      sites %>%
      dplyr::filter(gene == i) %>%
      dplyr::left_join(refs, by = c('chr', 'pos')) %>%
      tidyr::pivot_longer(cols = dplyr::all_of(nucleotides),
                          values_to = 'nt_n_reads', names_to = 'nt') %>%
      dplyr::mutate(
        strandless_nt = 
          ifelse(nt %in% c('-', '_'), 'DEL', toupper(nt)),
        is_ref = ifelse(strandless_nt == ref, 'ref', 'alt'),
        strandless_nt = paste0(strandless_nt, '_', is_ref) %>% 
          forcats::fct_relevel(unique(c(paste0(topupper(nucleotides), '_ref'),
                                        paste0(toupper(nucleotides), '_alt'))))) %>%
      ggplot(aes(x = pos, y = nt_n_reads, fill = strandless_nt)) + 
      geom_vline(aes(xintercept = mut_pos)) +
      geom_col(position = 'stack') +
      theme_bw() +
      theme(axis.text.x = element_blank()) +
      ggh4x::facet_grid2(celltype ~ mut_pos, scales = 'free') +
      scale_fill_manual(values = c(A_ref = 'darkgreen', A_alt = 'lightgreen',
                                   C_ref = 'darkblue', C_alt = 'lightblue',
                                   G_ref = 'darkorange3', G_alt = 'peachpuff',
                                   T_ref = 'darkred', T_alt = 'salmon',
                                   INS_alt = 'black', DEL_alt = 'grey')) +
      ggtitle(i)
    print(p)
    })
```
